\section{Manifolds}

\subsection{Smooth manifolds}

\datum{2024-10-1}

\begin{definicija}
A space $M$ is \emph{paracompact}\index{paracompact} if for every
open covering $\setb{U_j}{j \in I}$ there exists a locally finite
refinement. That is, every point $p \in M$ has a neighbourhood that
intersects only finitely many sets of the refinement.
\end{definicija}

\begin{opomba}
Instead of second-countable, we require manifolds to be
paracompact.
\end{opomba}

\begin{definicija}
The \emph{Grassmanian}\index{Grassmanian} is the space
\[
G_{k,n}(\C) =
\setb{V \leq \C^n}{\dim V = k}.
\]
\end{definicija}

\begin{trditev}
The Grassmanian is a complex manifold of dimension $k (n-k)$.
\end{trditev}

\begin{proof}
Let $M_{k,n}(\C)$ be the set of $k \times n$ matrices of maximal
rank. This is an open subspace of $\C^{k \times n}$, therefore it
is a manifold of dimension $k \cdot n$. We can then define a map
$\pi \colon M_{k,n}(\C) \to G_{k,n}(\C)$ mapping each matrix to the
span of its rows. Note that $\pi(A) = \pi(B)$ if and only if there
exists some $g \in \GL_k(\C)$ with $gA = B$. That is,
\[
G_{k,n}(\C) = \kvoc{M_{k,n}(\C)}{\GL_k(\C)}.
\]
For $A \in M_{k,n}(\C)$, denote by $A_1, \dots, A_\ell$ its
$k \times k$-minors. As $\rank A = k$, there exists an index $j$
such that $\det A_j \ne 0$. Let
\[
U_j = \setb{\pi(A)}{\det A_j \ne 0}.
\]
By the above observation, this is an open covering of
$G_{k,n}(\C)$. Define $\varphi_j \colon U_j \to \C^{k(n-k)}$ as
follows -- choose a matrix $A$ such that $A_j \ne 0$. Then define
\[
\varphi_j(\pi(A)) = A_j^{-1} \cdot B_j,
\]
where $B_j$ is $A$ without the columns of $A_j$. This is clearly a
homomorphism. It is easy to see that the transition maps are
holomorphic.
\end{proof}

\begin{definicija}
A smooth manifold is \emph{orientable}\index{orientable} if there
exists an atlas such that the jacobians of its transition maps have
positive determinants.
\end{definicija}

\datum{2024-10-8}

\begin{trditev}
Let $M$ be a complex manifold of complex dimension $n$. Then $M$ is
also a smooth orientable manifold of real dimension $2n$.
\end{trditev}

\begin{proof}
We can clearly see that $M$ is a smooth manifold of real dimension
$2n$. Let $\setb{(U_\alpha, \varphi_\alpha)}{\alpha \in I}$ be a
complex atlas for $M$. For $\alpha, \beta \in I$, take
$F = \varphi_\alpha \circ \varphi_\beta^{-1}$ and write
$F(z) = u(x,y) + i v(x,y)$, where $z = x + iy$ for $x, y \in \R^n$.
Then
\[
dF =
\begin{bmatrix}
\parc{u}{x} & \parc{u}{y} \\
\parc{v}{x} & \parc{v}{y}
\end{bmatrix}
=
\begin{bmatrix}
 \parc{u}{x} & \parc{u}{y} \\
-\parc{u}{y} & \parc{u}{x}
\end{bmatrix},
\]
therefore
\[
\det (dF) = \br{\det \parc{u}{x}}^2 + \br{\det \parc{u}{y}}^2 > 0.
\qedhere
\]
%TODO wtf?
\end{proof}

\newpage

\subsection{Structure sheaf, tangent bundle, and differentials}

\begin{definicija}
Let $M$ be a smooth manifold and $p \in M$.
Consider\footnote{Replace $\mathcal{C}^\infty$ with $\mathcal{O}$
for complex manifolds.}
\[
G_p =
\setb{(U, f)}
{p \in U \land f \in \mathcal{C}^\infty(U) \land 
\text{$U$ is open}}.
\]
We define an equivalence relation on $G_p$ by
$(U, f) \sim (V, g)$ if there exists an open set
$W \subseteq U \cap V$ containing $p$ such that
$\eval{f}{W}{} = \eval{g}{W}{}$. The set $\kvoc{G_p}{\sim}$ is the
space of \emph{germs}\index{germs} of functions at $p$, denoted by
$\mathcal{C}_{M,p}^\infty$. The equivalence class of $(U, f)$ is
denoted by $f_p$.
\end{definicija}

\begin{opomba}
The sets $\mathcal{C}_{M,p}^\infty$ and $\mathcal{O}_{M,p}$ are
commutative rings with a neutral element. The latter is also an
integral domain.
\end{opomba}

\begin{opomba}
The map $\mathcal{C}_{M,p}^\infty \to \R$, given by
$f_p \mapsto f(p)$, is well defined and the set
$\mathcal{M}_{M,p} = \setb{f_p}{f(p) = 0}$ is a maximal ideal.
\end{opomba}

\begin{definicija}
Let $U \subseteq M$ be an open set. Define
\[
\mathcal{C}_M^\infty(U) =
\setb{f \colon U \to \R}{\text{$f$ is smooth}}.
\]
The operator $\mathcal{C}_M^\infty$ is called the
\emph{structure sheaf}\index{structure sheaf} of $M$.
\end{definicija}

\begin{definicija}
A \emph{derivation}\index{derivation} $v$ on
$\mathcal{C}_{M,p}^\infty$ is a linear operator
$v \colon \mathcal{C}_{M,p}^\infty \to \R$ such that
\[
v(f_p g_p) = f(p) v(g_p) + g(p) v(f_p)
\]
for all $f_p, g_p \in \mathcal{C}_{M,p}^\infty$. The set of all
derivations on $\mathcal{C}_{M,p}^\infty$ is denoted by
$\Tang_p \! M$.
\end{definicija}

\begin{opomba}
Let $(U, \varphi)$ be a local chart around $p$ such that
$\varphi(p) = 0$. Then
$\varphi^* \colon
\mathcal{C}_{\R^n,0}^\infty \to \mathcal{C}_{M,p}^\infty$, given by
$\varphi^*(f) = f \circ \varphi$, is a ring isomorphism.
\end{opomba}

\begin{opomba}
The map $d \varphi_p \colon \Tang_p M \to \Tang_0 \R^n$, given
by
\[
d \varphi_p(v)(f) = v(f \circ \varphi),
\]
is an isomorphism of vector spaces.
\end{opomba}

\begin{definicija}
We define
\[
\parc{}{x_j^\alpha}(p)(f) =
\eval{\parc{\br{f \circ \varphi_\alpha^{-1}}}{x_j}}{0}{}
\]
for $f \in \mathcal{C}_{M,p}^\infty$.
\end{definicija}

\begin{lema}
The set
\[
\setb{\parc{}{x_j}(p)}{j \leq n}
\]
is a basis of $\Tang_p \! M$.
\end{lema}

\begin{proof}
We can assume $M=\R^n$ and $p=0$. Take
$f \in \mathcal{C}_{\R^n,0}^\infty$ and $v \in \Tang_0 \! \R^n$.
Then
\[
v(f) =
v \br{f(0) + \sum_{j=1}^n c_j x_j + O \br{\abs{x}^2}} =
\sum_{j=1}^n c_j v(x_j),
\]
since $v \br{O \br{\abs{x}^2}} = 0$. Hence
\[
v = \sum_{j=1}^n v(x_j) \parc{}{x_j}(p)
\]
as $c_j = \parc{f}{x_j}(0)$. It follows that the above set is in
fact a generator of $\Tang_p \! M$. As they are clearly linearly
independent, they indeed form a basis.
\end{proof}

\begin{definicija}
Let $M$ and $N$ be manifolds and $f \colon M \to N$ be a smooth
map. For $p \in M$, define
$d f_p \colon \Tang_p \! M \to \Tang_{f(p)} \! N$ via
\[
d f_p (v)(h) = v(h \circ f)
\]
for $v \in \Tang_p \! M$ and $h \in \mathcal{C}_{N,f(p)}^\infty$.
\end{definicija}

\begin{trditev}
Let $p \in M$ and $f \colon M \to N$ be a smooth map. Let
$(U, \varphi)$ be a local chart around $p$ and $(V, \psi)$ be a
local chart around $f(p)$. Then the matrix of the linear map
$d f_p$ with respect to the standard bases is given by the jacobian
matrix of $\psi \circ f \circ \varphi^{-1}$ in $\varphi(p)$.
\end{trditev}

\begin{definicija}
Let $M$ be a smooth manifold of dimension $n$. A topological
subspace $N \subseteq M$ is a
\emph{regular submanifold}\index{regular submanifold} of
codimension $k$ if for every point $p \in N$ there exists a local
chart $(U, \varphi)$ near $p \in M$ such that
\[
\varphi(N \cap U) =
\setb{x \in \varphi(U) \subseteq \R^n}
{\forall j \leq k \colon x_j = 0}.
\]
\end{definicija}

\begin{trditev}
If $N$ is a regular submanifold, then for every point $p \in N$
there exists a neighbourhood $U$ of $p$ and a smooth map
$F \colon U \to \R^k$ such that
$N \cap U = \setb{q \in U}{f(q) = 0}$ and $d F_q$ is surjective for
all $q \in N \cap U$.
\end{trditev}

\begin{proof}
Take $F$ to be the projection on the first $k$ coordinates.
\end{proof}

\begin{izrek}[Rank]
\index{rank theorem}
Let $M_1$ and $M_2$ be smooth manifolds of dimensions $n$ and $m$.
Let $F \colon M_1 \to M_2$ be a smooth map. Assume that
$\rank(d f_p) = k \in \N$ for all $p \in M_1$. Then for all
$a \in F(M_1)$ we have that $F^{-1}(a)$ is a regular submanifold
of dimension $k$ in $M_1$.
\end{izrek}

\begin{trditev}
Let $(a_0, \dots, a_n) \in \C^{n+1} \setminus \set{0}$ and define
\[
H =
\setb{[z_0 : \dots : z_n] \in \proj{\C}{n}}
{\sum_{j=0}^n a_j z_j = 0}.
\]
Then $H$ is a submanifold.
\end{trditev}

\newpage

\subsection{Coverings and group actions}

\begin{trditev}
Let $M$ be a smooth manifold with covering
$\br{\widetilde{M},\pi}$. Then $\widetilde{M}$ has a unique smooth
structure such that $\pi$ is smooth.
\end{trditev}

\begin{definicija}
A group $G$ is a \emph{Lie group}\index{Lie group} if it is a
topological space admitting a smooth structure such that the
operation $G \times G \to G$, given by $(g,h) \mapsto g h^{-1}$, is
smooth.
\end{definicija}

\begin{definicija}
For a smooth manifold $M$ denote
\[
\Aut(M) = \setb{f \colon M \to M}{\text{$f$ is a diffeomorphism}}.
\]
A group $G$ \emph{acts}\index{group action} on $M$ via
$\varphi \colon G \to \Aut(M)$ if $\varphi$ is a homomorphism. The
action is \emph{faithful}\index{faithful action} or
\emph{effective} if $\varphi$ is injective.
\end{definicija}

\begin{definicija}
Given an action $\varphi \colon G \to \Aut(M)$, we define the
\emph{orbit}\index{orbit} of $x \in M$ as
\[
Gx = \setb{\varphi(g)(x)}{g \in G}.
\]
Define an equivalence relation on $M$ as $x \sim y \iff Gx = Gy$.
Denote $\kvoc{M}{G} = \kvoc{M}{\sim}$.
\end{definicija}

\begin{opomba}
For a smooth manifold $M$, $\kvoc{M}{\Aut(M)}$ is a point.
\end{opomba}

\begin{definicija}
A \emph{complex torus}\index{complex torus} is the quotient
$\kvoc{\C^n}{G}$, where
\[
G =
\setb{z \mapsto z + \sum_{j=1}^{2n} \lambda_j w_j}
{\forall j \colon \lambda_j \in \Z}
\]
for $\R$-linearly independent vectors $w_j$.
\end{definicija}
