\section{Basic homotopy theory}

\subsection{Definition}

\datum{2023-10-2}

\begin{definicija}
Continuous maps $f, g \colon X \to Y$ of topological spaces are
\emph{homotopic}\index{homotopic}, if there is a continuous map
$H \colon X \times I \to Y$, such that $H(x, 0) = f(x)$ and
$H(x, 1) = g(x)$. Such $H$ is called a
\emph{homotopy}\index{homotopy}. We write $H \colon f \simeq g$.
\end{definicija}

\begin{opomba}
If $X$ is a locally compact and Hausdorff space, homotopies
coincide with paths in the space $\mathcal{C}(X, Y)$.
\end{opomba}

\begin{trditev}
Homotopy is an equivalence relation on $\mathcal{C}(X, Y)$.
\end{trditev}

\obvs

\begin{definicija}
We denote the set of equivalence classes of the homotopy relation
on $\mathcal{C}(X, Y)$ by $[X, Y]$.
\end{definicija}

\begin{opomba}
If $X$ is a locally compact and Hausdorff space, $[X, Y]$ is the
set of path components of $\mathcal{C}(X, Y)$.
\end{opomba}

\begin{definicija}
With $f \colon (X, A) \to (Y, B)$ we denote maps $f \colon X \to Y$
such that $f(A) \subseteq f(B)$. Similarly, we define
$\mathcal{C}( (X, A), (Y, B))$ and $[(X,A), (Y,B)]$.
\end{definicija}

\begin{definicija}
Let $A \subseteq X$ and $f, g \colon X \to Y$ be maps satisfying
$\eval{f}{A}{} = \eval{g}{A}{}$. The map
$G \colon X \times I \to Y$ is a
\emph{homotopy relative to $A$}\index{homotopy!relative to a subspace}
if $H \colon f \simeq g$ and $\eval{H_t}{A}{} = \eval{f}{A}{}$ for
all $t \in I$. We write $H \colon f \simeq g~(\rel A)$.
\end{definicija}

\datum{2023-10-6}

\begin{definicija}
A map $f \colon X \to Y$ is
\emph{null-homotopic}\index{null-homotopic} if it is homotopic to a
constant.
\end{definicija}

\begin{definicija}
Let $X$ be a topological space. The \emph{cone}\index{cone} on $X$
is the space
\[
CX = \kvoc{X \times I}{X \times \set{1}}.
\]
\end{definicija}

\begin{trditev}
A map $f \colon X \to Y$ is null-homotopic if and only if it
extends to the cone $CX$.
\[
\begin{tikzcd}[column sep=large, row sep=large]
X \arrow[r, "f"] \arrow[d, hook, "i"'] & Y \\
CX \arrow[ur, dashed, "F"']
\end{tikzcd}
\]
\end{trditev}

\begin{proof}
In the following diagram, $H$ exists if and only if $F$ exists.
\[
\begin{tikzcd}[column sep=large, row sep=large]
X \arrow[r, "f"] \arrow[d, hook, "i_0"'] & Y \\
X \times I \arrow[ur, dashed, "H"] \arrow[r, "q"']
&
CX \arrow[u, dashed, "F"']
\end{tikzcd} \qedhere
\]
\end{proof}

\newpage

\subsection{Homotopy equivalence}

\begin{trditev}
Let $f, f' \colon X \to Y$ and $g, g' \colon Y \to Z$ be maps. If
$f \simeq g$ and $f' \simeq g'$, we also have
$g \circ f \simeq g' \circ f'$.
\end{trditev}

\begin{proof}
Let $H \colon X \times I \to Y$ and $K \colon Y \times I \to Z$ be
the two homotopies. It is trivial to check that
\[
L(x, t) = K(H(x, t), t)
\]
is a homotopy of the compositions.
\end{proof}

\begin{definicija}
The \emph{homotopy category}\index{homotopy!category}
$\catt{HoTop}$ is the category with topological spaces as objects
and homotopies as morphisms. Operations are induced by the
compositions of maps.
\end{definicija}

\begin{definicija}
The category $\catt{Top}^2$ has pairs of spaces $(X, A)$ with
$A \subseteq X$ as spaces, maps of pairs as morphisms and
composition as the operation. Analogously as above, this category
induces the category $\catt{HoTop}^2$.
\end{definicija}

\begin{definicija}
The category $\catt{Top}_\bullet$ has pairs $(X, x_0)$ with
$x_0 \in X$ as spaces, maps of pairs as morphisms and composition
as the operation. Analogously as above, this category induces the
category $\catt{HoTop}_\bullet$.
\end{definicija}

\begin{definicija}
\emph{Homotopy equivalence}\index{homotopy!equivalence} is an
isomorphism in the category $\catt{HoTop}$. If spaces $X$ and $Y$
are homotopy equivalent, we write $X \simeq Y$.
\end{definicija}

\begin{opomba}
A map $f \colon X \to Y$ is a homotopy equivalence if there exists
a map $g \colon Y \to X$ such that both $g \circ f \simeq \id_X$
and $f \circ g \simeq \id_Y$ hold. The map $g$ is a
\emph{homotopy inverse}\index{homotopy!inverse} of $f$.
\end{opomba}

\begin{definicija}
A space $X$ is \emph{contractible}\index{contractible space} if
it is homotopy equivalent to a point.
\end{definicija}

\begin{opomba}
Every cone is contractible.
\end{opomba}

\begin{trditev}
Let $X$ be a topological space. The following statements are
equivalent:

\begin{enumerate}[i)]
\item The space $X$ is contractible.
\item The map $\id_X$ is homotopy equivalent to a constant map.
\item The space $X$ is a retract of $CX$.
\end{enumerate}
\end{trditev}

\obvs

\begin{izrek}
Let $X$ and $Y$ be closed surfaces. If $X \simeq Y$, then
$X \approx Y$.
\end{izrek}

\begin{definicija}
A subspace $A \subseteq X$ is a
\emph{deformation retract}\index{deformation retract} of $X$ if it
is a retract and the retraction is a homotopy inverse of the
inclusion.
\end{definicija}

\begin{definicija}
A subspace $A \subseteq X$ is a
\emph{strong deformation retract}\index{strong deformation retract}
if it is a retract and the retraction is a homotopy inverse of the
inclusion relative to $A$.
\end{definicija}

\newpage

\subsection{Extensions of homotopies}

\datum{2023-10-9}

\begin{definicija}
A closed subspace $A \subseteq X$ has the
\emph{homotopy extension property}\index{homotopy!extension property}
if for every space $Y$, map $f \colon X \to Y$ and homotopy
$H \colon A \times I \to Y$ with $H_0 = \eval{f}{A}{}$ there exists
a homotopy $\oline{H} \colon X \times I \to Y$ such that
$\oline{H}_0 = f$ and $\eval{\oline{H}}{A}{} = H$.
\end{definicija}

\begin{trditev}
A closed subspace $A \subseteq X$ has the homotopy extension
property if an only if the space
\[
L = A \times I \cup X \times \set{0}
\]
is a retract of $X \times I$.
\end{trditev}

\begin{proof}
Suppose that $L$ is a retract of $X$. It is easy to see that
$\oline{H} = (H \cup f) \circ r$ is the required homotopy
extension, where $r \colon X \times I \to L$ is a retraction.

Now suppose that $A$ has the homotopy extension property. Define
$i_0 \colon X \hookrightarrow L$ and
$H \colon A \times I \hookrightarrow L$ as the inclusions. By the
homotopy extension property, there exists a homotopy
$\oline{H} \colon X \times I \to L$, which is of course a
retraction.
\end{proof}

\begin{trditev}
Let $A \subseteq X$ be a contractible subspace. If $A$ has the
homotopy extension property, then $q \colon X \to \kvoc{X}{A}$ is a
homotopy equivalence.
\end{trditev}

\begin{proof}
Let $K \colon A \times I \to A$ be a homotopy equivalence between
$\id_A$ and the constant map. Then
$K \cup \id_X \colon A \times I \cup X \to X$ is a well defined
map. Let $H \colon X \times I \to X$ be its extension.
\[
\begin{tikzcd}[column sep=large, row sep=large]
X \times I \arrow[r, "H"] \arrow[d, "q \times \id_I"'] &
X \arrow[d, "q"] \\
\kvoc{X}{A} \times I \arrow[r, dashed, "\oline{H}"']
&
\kvoc{X}{A}
\end{tikzcd}
\]
The map $q \times \id_I$ is a quotient map because $I$ is locally
compact. The induced map $\oline{H}$ is therefore well defined.
But now $H_1$ induces a map $h \colon \kvoc{X}{A} \to X$. Note
that, by definition, $h \circ q = H_1 \simeq H_0 = \id_X$, and
$q \circ h = \oline{H}_1 \simeq \oline{H}_0 = \id_{\kvoc{X}{A}}$.
\end{proof}

\datum{2023-10-13}

\begin{trditev}
Suppose the map $A \hookrightarrow X$ has the homotopy extension
property and $f, g \colon A \to Y$ are homotopic. Then we have
$X \cup_f Y \simeq X \cup_g Y$.
\end{trditev}

\begin{trditev}
Suppose the map $A \hookrightarrow X$ has the homotopy extension
property. If $A \hookrightarrow X$ is a homotopy equivalence, then
$A$ is a strong deformation retract of $X$.
\end{trditev}

\begin{definicija}
Let $f \colon X \to Y$ be a map. The
\emph{mapping cylinder}\index{mapping cylinder} of $f$ is the space
\[
M_f = \kvoc{X \times I \sqcup Y}{(x,0) \sim f(x)}.
\]
\end{definicija}

\begin{opomba}
The subspace $X \approx X \times \set{1} \subseteq M_f$ has the
homotopy extension property.
\end{opomba}

\begin{trditev}
Suppose that $f \colon X \to Y$ is a homotopy equivalence. Then
both $X$ and $Y$ are strong deformation retracts of $M_f$.
\end{trditev}

\begin{proof}
Note that the map $H((x,s),t) = [(x, s(1-t)]$ is a homotopy between
$\id$ and a retraction $r \colon M_f \to Y$, therefore $Y$ is a
strong deformation retract of $M_f$. In particular, $r$ is a
homotopy equivalence. Denote by $i \colon X \hookrightarrow M_f$
the inclusion of $X \simeq X \times \set{1}$ in $M_f$. Note that
$f = r \circ i$, therefore $i$ is also a homotopy equivalence. It
follows that $X$ is also a strong deformation retract of $M_f$.
\end{proof}
