\section{Operators on Hilbert spaces}

\subsection{Matrices and bounded operators}

\datum{2024-2-21}

\begin{definicija}
Let $V$ be a finite-dimensional vector space over a field $K$. A
\emph{linear operator}\index{linear operator} $A$ in $V$ is a
linear map $A \colon V \to V$.
\end{definicija}

\begin{definicija}
Let $A \colon V \to V$ be a linear operator. A closed subspace
$U \leq V$ is an
\emph{invariant subspace}\index{invariant subspace} for $A$ if
$A(U) \subseteq U$. The set of all invariant subspaces of $A$ is
denoted by $\Lat(A)$.
\end{definicija}

\begin{opomba}
An operator $A$ is invariant for $U$ if we can write
\[
A =
\begin{bmatrix}
B & C \\
0 & D
\end{bmatrix}
\]
in the decomposition $V = U \oplus W$.
\end{opomba}

\begin{opomba}
If $U \in \Lat(A)$, then $p_{\eval{A}{U}{}} \mid p_A$.
\end{opomba}

\begin{definicija}
Let $q \in \N$ be the minimal integer such that\footnote{Such a $q$
exists as $V$ is finite-dimensional.}
\[
\ker (A - \lambda)^q = \ker (A - \lambda)^{q+1}.
\]
The subspace
$N_\lambda = \ker (A - \lambda)^q$ is called the
\emph{root subspace}\index{root subspace} of $A$.
\end{definicija}

\begin{definicija}
A subspace $U \leq V$ is a
\emph{cyclic subspace}\index{cyclic subspace} for
$A \colon V \to V$ if
\[
U = \spn \setb{A^n x}{0 \leq n \leq q}
\]
for some $x \in V$.
\end{definicija}

\begin{definicija}
Let $X$ be a Banach space. A linear operator $A \colon X \to X$ is
\emph{bounded}\index{bounded operator} if the set
\[
\setb{\frac{\norm{Ax}}{\norm{x}}}{x \in X \setminus \set{0}}
\]
is bounded.
\end{definicija}

\newpage

\subsection{Compact operators on Banach spaces}

\begin{definicija}
Let $X$ and $Y$ be Banach spaces. A linear operator
$T \colon X \to Y$ is \emph{compact}\index{compact operator} if $T$
maps bounded sets in $X$ into pre-compact sets in $Y$.
Equivalently, the set $T(B_X)$ is pre-compact.
\end{definicija}

\begin{trditev}
Let $k \in \mathcal{C}([0,1]^2)$ be a continuous function. Then the
integral operator
\[
(Kf)(x) = \int_0^1 k(x,y) f(y)\,dy
\]
is a compact operator on $(\mathcal{C}([0,1]), \norm{\cdot}_2)$ and
$(\mathcal{C}([0,1]), \norm{\cdot}_\infty)$.
\end{trditev}

\begin{proof}
Introduction to functional analysis, proposition~5.4.9.
\end{proof}

\begin{definicija}
A sequence $(x_n)_{n \in \N} \subseteq X$ is
\emph{weakly convergent}\index{weak convergence} with limit $x$ if
\[
\lim_{n \to \infty} f(x_n) = f(x)
\]
for all functionals $f \in X^*$.
\end{definicija}

\begin{definicija}
A sequence $(x_n)_{n \in \N} \subseteq X$ is
\emph{normally convergent}\index{normal convergence} with limit $x$
if
\[
\lim_{n \to \infty} \norm{x_n - x} = 0.
\]
\end{definicija}

\begin{izrek}
A compact operator maps weakly convergent sequences into normal
convergent sequences.
\end{izrek}

\begin{proof}
Let $T \colon X \to Y$ be a compact operator and let
$(x_n)_{n \in \N}$ be a weakly convergent sequence with limit $x$.
By the uniform boundedness principle,\footnote{Introduction to
functional analysis, theorem~3.3.5.} the sequence
$\br{\norm{x_n}}_{n \in \N}$ is bounded. But then for every
functional $f \in Y^*$ it holds that
\[
f(Tx_n) - f(Tx) = (T^* f)(x_n - x),
\]
hence $\br{Tx_n}_{n \in \N}$ is weakly convergent. Suppose that it
is not normally convergent -- that is, there exists some 
$\varepsilon > 0$ and a subsequence $\br{x_{n_k}}_{k \in \N}$ such
that
\[
\norm{Tx_{n_k} - Tx} \geq \varepsilon
\]
holds for all $k$. As $T$ is compact, this subsequence has an
accumulation point. The only possible accumulation point is
clearly $Tx$.
\end{proof}

\begin{opomba}
If $T$ is a bounded operator on a reflexive Banach space $X$, the
converse holds as well.
\end{opomba}

\begin{izrek}
If $(T_n)_{n \in \N}$ is a sequence of compact operator with
bounded limit $T$, then $T$ is compact.
\end{izrek}

\begin{proof}
Introduction to functional analysis, theorem~5.4.4.
\end{proof}

\begin{izrek}
Every compact operator on a Hilbert space is a normal limit of
finite rank operators.
\end{izrek}

\begin{proof}
Introduction to functional analysis, theorem~5.4.10.
\end{proof}
