\section{Approximation of holomorphic functions}

\epigraph{``Tomorrow we will look at the pole shifting lemma, which
allows us to move the poles away.''}
{-- viš.~znan.~sod.~dr.~Rafael Benedikt Andrist}

\subsection{Runge's little theorem}

\datum{2023-12-19}

\begin{lema}\label{app:lm:seg}
Let $\Omega \subseteq \C$ be an open subset and
$K \subseteq \Omega$ a non-empty compact. Then there exist finitely
many horizontal or vertical line segments
$\sigma_1, \dots, \sigma_n$ of equal length in
$\Omega \setminus K$ such that for all $f \in \mathcal{O}(\Omega)$
we have
\[
f(z) = \frac{1}{2 \pi i} \sum_{k=1}^n
\lint_{\sigma_k} \frac{f(\xi)}{\xi - z}\,d\xi
\]
for all $z \in K$.
\end{lema}

\begin{proof}
Define
\[
\delta =
\begin{cases}
          1,           & \Omega = \C,   \\
d(K, \partial \Omega), & \Omega \ne \C.
\end{cases}
\]
Let $Q$ be a grid of squares, parallel to the coordinate axes, with
side length $d < \frac{\delta}{\sqrt{2}}$. As $K$ is compact, it
only intersects finitely many of them. Now just choose the boundary
of the union of those squares. It is clear that all those segments
are subsets of $\Omega \setminus K$.

Let $Q_k$ denote the above squares. Then,
\[
\frac{1}{2 \pi i} \sum_{k=1}^m \olint_{\partial Q_k}
\frac{f(\xi)}{\xi - z}\,d\xi =
\frac{1}{2 \pi i} \sum_{k=1}^n \lint_{\sigma_k}
\frac{f(\xi)}{\xi - z}\,d\xi.
\]
If $z \in \Int Q_\ell$, then
\[
f(z) =
\frac{1}{2 \pi i} \lint_{\partial Q_\ell}
\frac{f(\xi)}{\xi - z}\,d\xi
\]
and
\[
\frac{1}{2 \pi i} \lint_{\partial Q_k}
\frac{f(\xi)}{\xi - z}\,d\xi = 0
\]
for all $k \ne \ell$, therefore the lemma holds for all such $z$.
As both sides of the equations are continuous functions, they must
agree on the whole set $K$.
\end{proof}

\begin{lema}\label{app:lm:app_seg}
Let $\sigma$ be a compact line segment and $K \subseteq \C$ be a
compact set such that $K \cap \sigma = \emptyset$. Let
$h \in \mathcal{C}(\sigma)$ be a function. Then for all
$\varepsilon > 0$ there exist points $c_1, \dots, c_m \in \C$ and
$w_1, \dots, w_m \in \sigma$ such that
\[
\norm{\lint_\sigma \frac{h(\xi)}{\xi - z}\,d\xi -
\sum_{k=1}^m \frac{c_k}{z - w_k}}_K <
\varepsilon.
\]
\end{lema}

\begin{proof}
Let $\ell$ be the length of $\sigma$ and define the function
$v \colon \sigma \times K \to \C$ with
\[
v(\xi, z) = \frac{h(\xi)}{\xi - z}.
\]
It is clearly continuous, therefore it is uniformly continuous. In
particular, there exists a $\delta > 0$ such that
\[
\abs{v(\xi, z) - v(\xi', z)} < \frac{\varepsilon}{\ell}
\]
for all $z \in \C$ and $\abs{\xi - \xi'} < \delta$.

Let $\tau_1, \dots, \tau_m$ be the partition of $\sigma$ into line
segments of length $d < \sigma$. Choose points $w_k \in \tau_k$ and
set $c_k = - h(w_k) d$. We therefore have
\[
\abs{\lint_{\tau_k} v(\xi, z)\,d\xi - \frac{c_k}{z - w_k}} =
\abs{\lint_{\tau_k} v(\xi, z)\,d\xi - d \cdot v(w_k, z)} <
d \cdot \frac{\varepsilon}{\ell}.
\]
Summing up, we get the desired inequality.
\end{proof}

\begin{lema}\label{app:lm:app_by_rat}
Let $\Omega \subseteq \C$ be an open set and $K \subseteq \Omega$
be a non-empty compact. Then there exist finitely many line
segments $\sigma_1, \dots, \sigma_n$ in $\Omega \setminus K$ such
that for any holomorphic function $f \in \mathcal{O}(\Omega)$ and
$\varepsilon > 0$ there exists a rational function $q$ of the form
\[
q(z) = \sum_{k=1}^m \frac{c_k}{z - w_k},
\]
where $c_k \in \C$ and $w_k \in \sigma$, such that
\[
\norm{f - q}_K < \varepsilon.
\]
\end{lema}

\begin{proof}
By lemma~\ref{app:lm:seg} there exist line segments $\sigma_k$ such
that
\[
f(z) = \frac{1}{2 \pi i} \sum_{k=1}^n
\lint_{\sigma_k} \frac{f(\xi)}{\xi - z}\,d\xi
\]
for all $z \in K$. Now just apply lemma~\ref{app:lm:app_seg} to
each line segment separately.
\end{proof}

\datum{2023-12-20}

\begin{lema}[Shifting poles]\index{Shifting poles lemma}
Let $K \subseteq \C$ be a compact set, $Z$ a connected component of
$\C \setminus K$ and $a, b \in Z$. Then, for all $\varepsilon > 0$
there exists a polynomial $q$ such that
\[
\norm{\frac{1}{z-a} - q \br{\frac{1}{z-b}}}_K < \varepsilon.
\]
If $Z$ is the unbounded component, we can approximate
$\frac{1}{z-a}$ by $q(z)$ instead.
\end{lema}

\begin{proof}
Let $L_\omega$ be the family of all functions that are holomorphic
in a neighbourhood of $K$ that can be approximated uniformly on $K$
by polynomials in $\frac{1}{z-\omega}$. Note that, if
$\frac{1}{z-p} \in L_q$, then $L_p \subseteq L_q$.

Consider the set
\[
S = \setb{s \in Z}{\frac{1}{z-s} \in L_b}.
\]
We claim that $S = Z$. First note that $b \in S$, therefore
$S \ne \emptyset$. Take any point $p \in S$ and $\delta > 0$ such
that $\dsk(p, \delta) \subseteq Z$. For any
$s \in \dsk(p, \delta)$, we can write
\[
\frac{1}{z-s} =
\frac{1}{z-p} \cdot \frac{1}{1 - \frac{s-p}{z-p}} =
\frac{1}{z-p} \cdot \sum_{k=0}^\infty
\br{\frac{s-p}{z-p}}^k,
\]
which converges as $\abs{s-p} < \abs{z-p}$ by choice of $\delta$.
We conclude that $\dsk(p, \delta) \subseteq S$. In particular, $S$
is an open set. Take any point $p \in \partial S \cap Z$. Suppose
that $\dsk(p, 3 \delta) \subseteq Z$ and choose a point
$p' \in \dsk(p, \delta) \cap S$. Then,
$\dsk(p', 2\delta) \subseteq S$ and therefore
$\dsk(p, \delta) \subseteq S$, which is a contradiction and
therefore proves our claim.

Now suppose that $Z$ is the unbounded component and take a point
$d \in Z$ such that $K \subseteq \dsk(0, \abs{d})$. Then all
functions $\br{\frac{1}{z-d}}^n$ can be approximated uniformly on
$K$ by Taylor's polynomials about $0$.
\end{proof}

\begin{definicija}
For any set $P \subseteq \C$ denote by $\C_P[z]$ the family of
rational functions with poles in $P$.
\end{definicija}

\begin{izrek}[Runge's approximation theorem]
\index{Runge's!theorem}
\label{app:thm:runge_app}
Let $K \subseteq \C$ be a compact. If $P$ intersects every bounded
connected component of $\C \setminus K$, then for every function
$f$, holomorphic in a neighbourhood of $K$, and every
$\varepsilon > 0$ there exists a function $q \in \C_P[z]$ such that
\[
\norm{f-q}_K < \varepsilon.
\]
\end{izrek}

\begin{proof}
Be lemma~\ref{app:lm:app_by_rat}, we can find a compact union
$\sigma$ of line segments such that every such $f$ can be
approximated by
\[
\widetilde{q} = \sum_{k=1}^m \frac{c_k}{z - w_k},
\]
where $c_k \in \C$ and $w_k \in \sigma$. Suppose then
\[
\norm{f - \widetilde{q}}_K < \frac{\varepsilon}{2}.
\]
Let $Z_k$ be the connected component of $\C \setminus K$ that
contains $w_k$.

If $Z_k$ is bounded, then choose $t_k \in P \cap Z_k$. By the pole
shifting lemma, we can find a polynomial $g_k$ such that
\[
\norm{\frac{c_k}{z - w_k} - g_k \br{\frac{1}{z - t_k}}}_K <
\frac{\varepsilon}{2m}.
\]
If $Z_k$ is unbounded, we can instead approximate
$\frac{c_k}{z - w_k}$ by a polynomial instead. Choosing
\[
q = \sum_{k=1}^m g_k,
\]
we find
\[
\norm{f - q}_K \leq
\norm{f - \widetilde{q}}_K + \norm{\widetilde{q} - q}_K <
\frac{\varepsilon}{2} + m \cdot \frac{\varepsilon}{2m} =
\varepsilon. \qedhere
\]
\end{proof}

\begin{izrek}[Runge's approximation theorem]
\index{Runge's!theorem}
Let $\Omega \subseteq \C$ be an open subset and let
$K \subseteq \Omega$ be a compact. If every bounded component of
$\C \setminus K$ intersects $\C \setminus \Omega$, then for every
function $f$ that is holomorphic on a neighbourhood of $K$ and
every $\varepsilon > 0$ there exists a function
$q \in \mathcal{O}(\Omega)$ such that
\[
\norm{f - q}_K < \varepsilon.
\]
\end{izrek}

\begin{proof}
Choose $P = \C \setminus \Omega$ in
theorem~\ref{app:thm:runge_app}.
\end{proof}

\begin{posledica}[Runge's little theorem]
\index{Runge's!little theorem}
Let $K \subseteq \C$ be a compact set such that $\C \setminus K$ is
connected. Then for every holomorphic function on a neighbourhood
of $K$ and $\varepsilon > 0$ there exists a polynomial
$q \in \C[z]$ such that
\[
\norm{f - q}_K < \varepsilon.
\]
\end{posledica}

\begin{proof}
Choose $P = \emptyset$ in theorem~\ref{app:thm:runge_app}.
\end{proof}

\newpage

\subsection{Applications of Runge's little theorem}

\datum{2024-1-3}

\begin{definicija}
Let $V$ be a vector space over $\C$, equipped with a topology, and
let $T \colon V \to V$ be a linear map.

\begin{enumerate}[i)]
\item We call $T$ \emph{cyclic}\index{cyclic!map} if there exists
some $f \in V$, called a \emph{cyclic vector}\index{cyclic!vector},
such that
\[
\spn_\C \setb{T^n(f)}{n \in \N_0} = V.
\]
\item We call $T$ \emph{hypercyclic}\index{hypercyclic map} if
there exists some $f \in V$, such that
\[
\oline{\setb{T^n(f)}{n \in \N_0}} = V.
\]
\end{enumerate}
\end{definicija}

\begin{izrek}[Birkhoff]
\index{Birkhoff's theorem}
Let $\tau \colon \C \to \C$ be given by $\tau(z) = z + a$ for some
$a \ne 0$. Then the map
$T \colon \mathcal{O}(\C) \to \mathcal{O}(\C)$, given by
$T(f) = f \circ \tau$, is hypercyclic.
\end{izrek}

\begin{proof}
Set $K_n = \oline{\dsk(\ell_n \cdot a, n)}$ for a sequence
$(\ell_n)_n \subseteq \N$, such that all $K_n$ are pairwise
disjoint and $K_n \cap \dsk(0, n) = \emptyset$ for all $n \in \N$.
Choose a sequence $(\varepsilon_n)_n \subseteq \R^+$, converging to
$0$. Furthermore, let $(p_n)_n$ be a sequence of all polynomials
$(\Q \oplus i\Q)[z]$.

We first construct a sequence of holomorphic functions $(f_n)_n$,
such that the following conditions hold:

\begin{enumerate}[i)]
\item For all $n \in \N$, we have
\[
\norm{\sum_{k=1}^n f_k - \tau^{-\ell_n} \circ p_n}_{K_n} <
\frac{\varepsilon_n}{2^n}.
\]
\item For all $m < n$, we have
\[
\norm{f_n}_{K_m} < \frac{\varepsilon_m}{2^n}.
\]
\item For all $n \in \N$, we have
\[
\norm{f_n}_{\dsk(0,n)} < \frac{1}{2^n}.
\]
\end{enumerate}

Choose $\ell_n$ inductively such that $K_n \cap K_m = \emptyset$
for all $m < n$. Then the set
\[
\C \bigsetminus \br{\bigcup_{m=1}^n K_m \cup \dsk(0,n)}
\]
is obviously connected. We can now just apply Runge's little
theorem by choosing the function $0$ on $\dsk(0,n)$ and $K_m$ for
$m < n$, and
\[
\tau^{-\ell_n} \circ p_n - \sum_{k=1}^{n-1} f_k
\]
on $K_n$.

Note that the series
\[
f = \sum_{n=1}^\infty f_n
\]
is uniformly convergent by the third condition. We now compute
\begin{align*}
\norm{f \circ \tau^{\ell_n} - p_n}_{\tau^{-\ell_n}(K_n)} &=
\norm{f - \tau^{-\ell_n} \circ p_n}_{K_n}
\\
&\leq
\sum_{k > n} \norm{f_k}_{K_n} +
\norm{\sum_{k=1}^n f_k - \tau^{-\ell_n} \circ p_n}_{K_n}
\\
&<
\varepsilon_n.
\end{align*}
It follows that we can approximate every rational polynomial with
iterations $T^n(f)$.
\end{proof}

\datum{2024-1-9}

\begin{definicija}
Let $X$ and $Y$ be topological spaces and $f \colon X \to Y$ a map.
We call $f$ \emph{proper}\index{proper map} if for every compact
set $K \subseteq Y$ its preimage $f^{-1}(K)$ is also compact.
\end{definicija}

\begin{definicija}
Let $\Omega \subseteq \C$ be an open set. A map
$f \colon \Omega \to \C^n$  is
\emph{holomorphic}\index{holomorphic} if every component is
holomorphic.
\end{definicija}

\begin{opomba}
If $\Omega \subseteq \C$ is a bounded domain, a continuous map
$f \colon \Omega \to \C^n$ is proper if and only if for every
sequence $(z_n)_n$ with limit $z \in \partial \Omega$ we have
\[
\lim_{n \to \infty} \norm{f(z_n)} = \infty.
\]
\end{opomba}

\begin{definicija}
A \emph{horseshoe compact}\index{horseshoe compact} is the set
\[
A(r,R) \setminus T,
\]
where $A(r,R)$ is a closed annulus with $r < R < 1$ and $T$ is an
open triangle with vertices $1$, $i \varepsilon$ and
$-i \varepsilon$.
\end{definicija}

\begin{lema}
Let $K_n$ be a sequence of horseshoe compacts that approximate the
boundary of $\dsk$ away from $1$. Then there exists some
$f \in \mathcal{O}(\dsk)$ such that
\[
\min_{z \in K_n} \abs{f(z)} \geq 2^n.
\]

\begin{figure}[!ht]
\centering
\begin{asy}
size(7cm);
real EPS = 0.2;
path horseshoe(real r, real R){
	pair X = intersectionpoint(circle(0,r),(0,EPS)--(1,0));
	pair Y = intersectionpoint(circle(0,R),(0,EPS)--(1,0));
	pair Z = intersectionpoint(circle(0,R),(0,-EPS)--(1,0));
	pair W = intersectionpoint(circle(0,r),(0,-EPS)--(1,0));
	return arc(0,Y,Z)--reverse(arc(0,X,W))--cycle;
}
draw(circle(0,1));
label(Label("$\dsk$", Relative(0.125)), circle(0,1));
path[] p;
real[] radii = {0.3, 0.5, 0.6, 0.75, 0.8, 0.9, 0.94, 0.97};
for(int t = 0; t < radii.length - 1; t += 2){
	p = p ^^ horseshoe(radii[t], radii[t+1]);
}
filldraw(p, cyan+opacity(0.2), blue);
\end{asy}
\caption{A sequence of horseshoe compacts}
\end{figure}
\end{lema}

\begin{proof}
Choose a sequence of disks $(V_n)_n$ such that
\[
\bigcup_{i=1}^{n-1} K_i \subset V_n,
\]
$V_{n-1} \subset V_n \subset \dsk$, $K_n \cap V_n \ne \emptyset$
and
\[
\bigcup_{n=1}^\infty V_n = \dsk.
\]
We now inductively construct a sequence of polynomials. Choose
$p_0(z) = 69$. For $p_n$, we choose a polynomial with
\[
\norm{p_n}_{V_n} < 2^{-n}
\quad \text{and} \quad
\min_{z \in K_n} \br{\abs{\sum_{i=0}^n p_i(z)}} \geq 2^n + 1.
\]
This can be done by Runge's little theorem. We can now take
\[
f(z) = \sum_{n=0}^\infty p_n(z),
\]
which converges normally on $\dsk$. We indeed have
\[
\abs{f(z)} \geq
\abs{\sum_{i=0}^n p_i(z)} - \sum_{i=n+1}^\infty \abs{p_i(z)} \geq
2^n
\]
for $z \in K_n$.
\end{proof}

\begin{izrek}
There exists a proper holomorphic embedding\footnote{Injective
immersion.} $F \colon \dsk \to \C^3$.
\end{izrek}

\begin{proof}
Take
\[
F(z) = \br{f(z), g(z), \frac{1}{1-z}},
\]
where $f$ and $g$ are as in the above lemma for alternating
sequences of horseshoe compacts. This map is clearly proper.
\end{proof}

\begin{opomba}
We can holomorphically embed the unit disk into $\C^2$.
\end{opomba}

\begin{izrek}
There exists a sequence $(p_n)_n$ of polynomials such that the
following conditions hold:

\begin{enumerate}[i)]
\item We have
\[
\lim_{n \to \infty} p_n(z) =
\begin{cases}
1, & z  =  0, \\
0, & z \ne 0.
\end{cases}
\]
\item For all $k \geq 1$ and $z \in \C$, we have
\[
\lim_{n \to \infty} p_n^{(k)}(z) = 0.
\]
\item The sequence $\br{p_n^{(k)}}_n$ converges uniformly on
compacts of $\C \setminus \R_{\geq 0}$, but does not converge
uniformly on any neighbourhood of any point in $\R_{\geq 0}$.
\end{enumerate}
\end{izrek}

\begin{proof}
Define
\[
I_n =
\setb{z \in \oline{\dsk \br{n}}}
{d \br{z, \R_{\geq 0}} \geq \frac{1}{n}}
\]
and
\[
K_n = \set{0} \cup \left[\frac{1}{n}, n \right] \cup I_n.
\]
Choose disjoint rectangular neighbourhoods $R_n$ of $0$ and $S_n$
of $\left[\frac{1}{n}, n \right]$, which are also disjoint with
$I_{n+1}$. Now define
\[
g_n(z) =
\begin{cases}
1, & z \in R_n,              \\
0, & z \in S_n \cup I_{n+1}.
\end{cases}
\]
As $\C \setminus \br{I_n \cup R_n \cup S_n}$ is connected, we can
apply Runge's little theorem to find a polynomial $p_n \in \C[z]$
with
\[
\norm{p_n - q_n}_{I_n \cup R_n \cup S_n} < \frac{1}{n^n}.
\]
It is clear that such a sequence of polynomials satisfies the first
condition. For any $k > 0$, Cauchy estimates give us
\[
\norm{p_n^{(k)}}_{I_{n-1}} =
\norm{p_n^{(k)} - q_n^{(k)}}_{I_{n-1}} \leq
\norm{p_n - q_n}_{I_n} \cdot
\frac{k!}{\br{\frac{1}{n(n-1)}}^k} \leq
\frac{k!}{n^n} \cdot (n(n-1))^k,
\]
which converges to $0$. The second condition is hence satisfied as
well. As $(I_n)_n$ exhausts $\C \setminus \R_{\geq 0}$, the
sequences of derivatives converge uniformly on this set.

Suppose now that $\br{p_n^{(k)}}_n$ converges uniformly on some
neighbourhood $U$ of a point in $\R_{\geq 0}$. It follows
(inductively) that the sequence $\br{p_n}_n$ converges uniformly on
$U$ as well. If $0 \in U$, this is clearly a contradiction, as the
limit function is not even continuous. Otherwise, let
$z_0 \in U \cap \R^+$. It is clear that the sequence $\br{p_n}_n$
converges uniformly on $\dsk(\abs{z_0})$. By the maximum principle,
it converges uniformly on $\dsk(\abs{z_0})$ as well, which is again
a contradiction.
\end{proof}

\begin{definicija}
Let $\Omega \subseteq \C$ be an open set and $K \subseteq \Omega$
be a compact. The
\emph{$\mathcal{O}(\Omega)$-convex hull}\index{O(Omega)-convex hull@$\mathcal{O}(\Omega)$-convex hull}
of $K$ is the set
\[
\widehat{K}_{\mathcal{O}(\Omega)} =
\setb{z \in \Omega}
{\forall f \in \mathcal{O}(\Omega) \colon
\abs{f(z)} \leq \norm{f}_K}.
\]
\end{definicija}

\datum{2024-1-10}

\begin{lema}
Let $\Omega \subseteq \C$ be an open subset and
$K \subseteq \Omega$ a compact subset.

\begin{enumerate}[i)]
\item For every component $Z$ of $\Omega \setminus K$ we have
$\Omega \cap \partial{Z} \subseteq K$. If $Z$ is relatively compact
in $\Omega$, then $\norm{f}_Z \leq \norm{f}_K$ for all
$f \in \mathcal{O}(\Omega)$.
\item Every component $Z_0$ of $\C \setminus K$ that is contained
in $\Omega$ is a component of $\Omega \setminus K$. If $Z_0$ is
bounded, then $Z_0 \subseteq \Omega$ is relatively compact.
\end{enumerate}
\end{lema}

\begin{proof}
Let $p$ be a point in $\Omega \cap \partial Z$ that is not in $K$.
Then there exists a disk $D$, centered at $p$, that is contained in
$\Omega \setminus K$. Note that every point of $D$ is in the same
connected components as $p$, which implies $p \in \Int Z$, a
contradiction.

Suppose now that $Z$ is relatively compact. As
$\partial Z \subseteq K$, we must have
\[
\norm{f}_Z = \norm{f}_{\partial Z} \leq \norm{f}_K
\]
by the maximum modulus principle.

For the second item, let $Z_0$ be a component of $\C \setminus K$
contained in $\Omega$. There must exist a connected component
$Z_1 \subseteq \Omega \setminus K$ that contains $Z_0$. Since
$Z_1$ is also a connected domain in $\C \setminus K$, it follows
that $Z_0 = Z_1$.

Finally, suppose that $Z_0$ is bounded. The closure of $Z_0$ in
$\C$ is contained in $\Omega$ by the first item. But then it is
compact as it is bounded and closed in $\Omega$.
\end{proof}

\begin{izrek}
Let $\Omega \subseteq \C$ be an open subset and
$K \subseteq \Omega$ a compact subset. The following statements are
equivalent:

\begin{enumerate}[i)]
\item The set $\Omega \setminus K$ has no relatively compact
components.\label{app:eqthm:1}
\item Every bounded component of $\C \setminus K$ intersects
$\C \setminus \Omega$.\label{app:eqthm:2}
\item Every function, holomorphic in a neighbourhood of $K$, can be
approximated uniformly on $K$ by rational functions without poles
in $\Omega$.\label{app:eqthm:3}
\item Every function, holomorphic in a neighbourhood of $K$, can be
approximated uniformly by functions, holomorphic in
$\Omega$.\label{app:eqthm:4}
\item For every point $p \in \Omega \setminus K$ there exists a
function $f \in \mathcal{O}(\Omega)$ such that
$\abs{f(p)} > \norm{f}_K$.\label{app:eqthm:5}
\item We have
$\widehat{K}_{\mathcal{O}(\Omega)} = K$.\label{app:eqthm:6}
\end{enumerate}
\end{izrek}

\datum{2024-1-16}

\begin{proof}
As usual, we will prove a sequence of implications.

Suppose first that $\Omega \setminus K$ has no relatively compact
components. If $Z_0$ is a bounded component of $\C \setminus K$
that is contained in $\Omega$, then it is also a component of
$\Omega \setminus K$ and relatively compact in $\Omega$ by the
previous lemma. This is a clear contradiction,
therefore~\ref{app:eqthm:1} implies~\ref{app:eqthm:2}.
    
From theorem~\ref{app:thm:runge_app} we immediately get that
statement~\ref{app:eqthm:2} implies~\ref{app:eqthm:3}. Trivially,
we have that~\ref{app:eqthm:3} implies~\ref{app:eqthm:4}.

Suppose now that~\ref{app:eqthm:4} holds and that the set
$\Omega \setminus K$ has a relatively compact component $Z$. Choose
a point $a \in Z$ and consider the function
$f(z) = \frac{1}{z - a}$. By our assumption, there exists a
function $g \in \mathcal{O}(\Omega)$ such that
\[
\norm{\frac{1}{z - a} - g}_K < \frac{1}{\norm{z-a}_K}.
\]
Then $\norm{1 - (z - a)g}_K < 1$. By the previous lemma, we
conclude
\[
\norm{1 - (z - a)g}_Z \leq \norm{1 - (z - a)g}_K < 1,
\]
but this is not the case for $z = a$. This
implies~\ref{app:eqthm:1}.

As~\ref{app:eqthm:5} and~\ref{app:eqthm:6} are equivalent by
definition, it remains to check that~\ref{app:eqthm:5} is
equivalent to~\ref{app:eqthm:1}.

Suppose that~\ref{app:eqthm:5} holds and that $Z$ is a relatively
compact component of $\Omega \setminus K$. As before, we have that
$\norm{f}_Z \leq \norm{f}_K$ for all functions
$f \in \mathcal{O}(\Omega)$. In particular, this holds for $f$ from
the assumption of statement~\ref{app:eqthm:5}, which is again a
contradiction.

Finally, suppose that $\Omega \setminus K$ has no relatively
compact components and let $p \in \Omega \setminus K$ be a point.
Consider the set $\Omega' = \Omega \setminus (K \cup \set{p})$,
which has the same connected components as $\Omega \setminus K$,
and are therefore still not relatively compact.\footnote{We did
remove a point from one of the components, but this property
remains unchanged.} From the previous arguments, it follows
that~\ref{app:eqthm:4} holds for $\Omega'$. In particular, let $g$
be a holomorphic function, defined on a neighbourhood of
$K \cup \set{p}$, with $\eval{g}{K}{} = 0$ and $g(p) = 1$. Thus
there exist a function $h \in \mathcal{O}(\Omega)$ such that
\[
\norm{g - h}_{K \cup p} < \frac{1}{2}.
\]
Then $\norm{h}_K < \frac{1}{2} < \abs{h(p)}$, which is exactly the
statement~\ref{app:eqthm:5}.
\end{proof}

\begin{lema}
Let $\Omega \subseteq \C$ be an open subset and
$K \subseteq \Omega$ a compact subset. Then there exists a compact
set $K'$ with $K \subseteq K' \subseteq \Omega$, such that every
bounded component of $\C \setminus K'$ intersects
$\C \setminus \Omega$ and contains a compact component of
$\C \setminus \Omega$.
\end{lema}

\begin{proof}
For the case $\Omega = \C$, we can take $K' = \oline{\dsk(0,r)}$,
where $r$ is large enough so that $K \subseteq K'$. Suppose then
that $\Omega \ne \C$. Choose
$0 < \rho < d(K, \partial \Omega) = d(K, \C \setminus \Omega)$ and
let
\[
M = \setb{z \in \Omega}{d(z, \C \setminus \Omega) \geq \rho}.
\]
Choose $r > 0$ such that $K \subseteq \oline{\dsk(0,r)}$ and set
$K' = M \cap \oline{\dsk(0,r)}$, which is clearly compact,
contained inside $\Omega$, and contains $K$. It remains to prove
the statement about the bounded components.

Let $Z$ be a bounded component of $\C \setminus K'$. Note that
$\C \setminus K' =
\br{\C \setminus M} \cup \br{\C \setminus \oline{\dsk(0,r)}}$. As
$Z$ is bounded, it must then be contained in $\C \setminus M$, as
$\C \setminus \oline{\dsk(0,r)}$ is connected and unbounded.

We can rewrite
\[
\C \setminus M = \bigcup_{w \in \C \setminus \Omega} \dsk(w, \rho)
\]
by the definition of $M$. For each such disk, we either have
$\dsk(w, \rho) \subseteq Z$ or $\dsk(w, \rho) \cap Z = \emptyset$.
Thus we can write
\[
Z = \bigcup_{w \in Z \setminus \Omega} \dsk(w, \rho).
\]
In particular, $Z$ intersects $\C \setminus \Omega$. Let $S$ be the
component of $\C \setminus \Omega$ that $Z$ intersects. As
$\C \setminus \Omega \subseteq \C \setminus K'$, $S$ is also
connected in $\C \setminus K'$. As $Z$ is a maximal connected
subset, we conclude that $S \subseteq Z$, as desired.
\end{proof}

\begin{definicija}
Let $\Omega \subseteq \C$ be an open subset. We call every bounded
component of $\C \setminus \Omega$ a \emph{hole}\index{hole} of
$\Omega$.
\end{definicija}

\begin{opomba}
If $K$ is the Cantor set, then $\Omega = \C \setminus K$ has
uncountably many holes.
\end{opomba}

\begin{izrek}[Runge's theorem for rational approximations]
\index{Runge's!theorem}
Let $\Omega \subseteq \C$ be an open set and
$P \subseteq \C \setminus \Omega$ a set such that $\oline{P}$
intersects every hole of $\Omega$. Then $\C_P[z]$ is dense in
$\mathcal{O}(\Omega)$.
\end{izrek}

\begin{proof}
Let $K \subseteq \Omega$ be compact subset. Construct $K'$ as in
the previous lemma. Then $\oline{P}$ intersects every bounded
component of $\C \setminus K'$. We can thus apply
theorem~\ref{app:thm:runge_app} for the compact $K'$.
\end{proof}

\begin{posledica}[Runge's theorem for polynomial approximation]
\index{Runge's!theorem}
If $\Omega$ has no holes then $\C[z]$ is dense in
$\mathcal{O}(\Omega)$.
\end{posledica}

\begin{proof}
Take $P = \emptyset$.
\end{proof}

\begin{izrek}[Mittag-Leffler for open subsets]
\index{Mittag-Leffler!theorem}
Let $\Omega \subseteq \C$ be an open subset. Let $(a_k, q_k)_k$ be
a distribution of principal parts in $\Omega$ and
$A = \setb{a_k}{k \in \N}$. Then there exists a Mittag-Leffler
series for $(a_k, q_k)$ that converges normally in
$\C \setminus A'$ where $A' = \oline{A} \setminus A$.
\end{izrek}

\begin{proof}
This is the same theorem as~\ref{inf_prod:lm:MLTOpen}. We present
another proof using Runge's approximation theorem.

First note that $A' \subseteq \partial \Omega$. Let 
\[
K_j =
\oline{\setb{z \in \Omega}
{\abs{z} \leq j \land d(z, \partial \Omega) \geq \frac{1}{j}}},
\]
which form an exhaustion by compacts. Let $K_j'$ be the compact
furnished by the previous lemma for $K_j$. Choose a subsequence
$(L_j)_j$ of $(K_j')_j$ that is an exhaustion by compacts, which
can be done by the properties of exhaustion, i.e.~for every $K_j'$
there exists some $k \in \N$ so that
$K_j' \subseteq K_k \subseteq K_k'$.

We may assume $K_1 \cap \oline{A} = \emptyset$. Let
$A_n = A \cap (L_{n + 1} \setminus L_n)$. For each $a \in A_n$ the
principal part of $q$ is holomorphic in $L_n$. By Runge's theorem,
there exists a function $g \in \mathcal{O}(\Omega)$ such that
\[
\norm{q - g}_{L_n} < \frac{1}{2^n \cdot \abs{A_n}}.
\]
Then
\[
\sum_{n=1}^\infty \norm{q_n - g_n}_{L_n} \leq
\sum_{n=1}^\infty 2^{-n} = 1,
\]
thus the series converges uniformly on compacts, as each compact
subset is contained in all but finitely many $L_n$.
\end{proof}

\begin{definicija}
A pair of sets $(\Omega, \Omega')$ is a
\emph{Runge pair}\index{Runge pair} if for every function
$f \in \mathcal{O}(\Omega)$, $\varepsilon > 0$ and compact
$K \subseteq \Omega$ there exists a function
$F \in \mathcal{O}(\Omega')$ such that
\[
\norm{f - F}_K < \varepsilon.
\]
\end{definicija}

\datum{2024-1-17}

\begin{izrek}[Šura-Bura]
Every compact connected component $K$ of a locally compact
Hausdorff space $X$ has a neighbourhood basis consisting of open
compact sets in $X$.
\end{izrek}

\begin{lema}Let $\Omega \subseteq \Omega' \subseteq \C$ be open
sets.

\begin{enumerate}[i)]
\item Every component of $Z$ of $\C \setminus \Omega$ with
$Z \subset \Omega'$ is a component of $\Omega' \setminus \Omega$.
\item For every open compact set
$K \subseteq \Omega' \setminus \Omega$ there exists a relatively
compact open set $V \subseteq \Omega'$ such that $K \subseteq V$
and $\partial V \subseteq \Omega$.
\end{enumerate}
\end{lema}

\begin{izrek}
Let $\Omega \subseteq \Omega' \subseteq \C$ be open sets. The
following statements are equivalent:

\begin{enumerate}[i)]
\item The set $\Omega' \setminus \Omega$ has no compact components.
\item The set $\C_P[z]$ with $P = \C \setminus \Omega'$ is dense in
$\mathcal{O}(\Omega)$.
\item The pair $(\Omega, \Omega')$ is a Runge pair.
\item The set $\Omega' \setminus \Omega$ contains no non-empty open
compact set.
\end{enumerate}
\end{izrek}

\begin{proof}
Suppose first that $\Omega' \setminus \Omega$ has no compact
components and set
$P = \C \setminus \Omega' \subseteq \C \setminus \Omega$. If
$Z \subseteq \C \setminus \Omega$ with $Z \subseteq \Omega'$ is a
component, it is also a component of $\Omega' \setminus \Omega$ by
the previous lemma. The set $Z$ therefore can't be compact. It
follows that every component either intersects $P$ or is unbounded.
By Runge's theorem for rational approximations, the set $\C_P[z]$
is dense in $\mathcal{O}(\Omega)$.

Note that the second statement implies the third by definition.

Suppose now that $(\Omega, \Omega')$ is a Runge pair and consider
an open compact set $K \subseteq \Omega' \setminus \Omega$ and
$a \in K$. Take the function $z \mapsto \frac{1}{z-a}$. By
definition of a Runge pair, we can find a sequence of holomorphic
functions $(g_n)_n \subseteq \mathcal{O}(\Omega')$ such that
\[
\lim_{n \to \infty} \norm{g_n - \frac{1}{z-a}}_{\partial V} = 0,
\]
where $V$ is chosen as in the above lemma. As $\partial V$ is
bounded, this implies
\[
\lim_{n \to \infty} \norm{g_n(z) \cdot (z-a) - 1}_{\partial V} = 0.
\]
By the maximum principle, we actually have
\[
\lim_{n \to \infty} \norm{g_n(z) \cdot (z-a) - 1}_V = 0,
\]
which gives a contradiction for $z=a$.

The last implication follows from Šura-Bura's theorem.
\end{proof}
