\section{Approximations of holomorphic functions}

\subsection{Runge's little theorem}

\datum{2023-12-19}

\begin{lema}\label{app:lm:seg}
Let $\Omega \subseteq \C$ be an open subset and
$K \subseteq \Omega$ a non-empty compact. Then there exist finitely
many horizontal or vertical line segments
$\sigma_1, \dots, \sigma_n$ of equal length in
$\Omega \setminus K$ such that for all $f \in \mathcal{O}(\Omega)$
we have
\[
f(z) = \frac{1}{2 \pi i} \sum_{k=1}^n
\lint_{\sigma_k} \frac{f(\xi)}{\xi - z}\,d\xi
\]
for all $z \in K$.
\end{lema}

\begin{proof}
Define
\[
\delta =
\begin{cases}
          1,           & \Omega = \C,   \\
d(K, \partial \Omega), & \Omega \ne \C.
\end{cases}
\]
Let $Q$ be a grid of squares, parallel to the coordinate axes, with
side length $d < \frac{\delta}{\sqrt{2}}$. As $K$ is compact, it
only intersects finitely many of them. Now just choose the boundary
of the union of those squares. It is clear that all those segments
are subsets of $\Omega \setminus K$.

Let $Q_k$ denote the above squares. Then,
\[
\frac{1}{2 \pi i} \sum_{k=1}^m \olint_{\partial Q_k}
\frac{f(\xi)}{\xi - z}\,d\xi =
\frac{1}{2 \pi i} \sum_{k=1}^n \lint_{\sigma_k}
\frac{f(\xi)}{\xi - z}\,d\xi.
\]
If $z \in \Int Q_\ell$, then
\[
f(z) =
\frac{1}{2 \pi i} \lint_{\partial Q_\ell}
\frac{f(\xi)}{\xi - z}\,d\xi
\]
and
\[
\frac{1}{2 \pi i} \lint_{\partial Q_k}
\frac{f(\xi)}{\xi - z}\,d\xi = 0
\]
for all $k \ne \ell$, therefore the lemma holds for all such $z$.
As both sides of the equations are continuous functions, they must
agree on the whole set $K$.
\end{proof}

\begin{lema}\label{app:lm:app_seg}
Let $\sigma$ be a compact line segment and $K \subseteq \C$ be a
compact set such that $K \cap \sigma = \emptyset$. Let
$h \in \mathcal{C}(\sigma)$ be a function. Then for all
$\varepsilon > 0$ there exist points $c_1, \dots, c_m \in \C$ and
$w_1, \dots, w_m \in \sigma$ such that
\[
\norm{\lint_\sigma \frac{h(\xi)}{\xi - z}\,d\xi -
\sum_{k=1}^m \frac{c_k}{z - w_k}}_K <
\varepsilon.
\]
\end{lema}

\begin{proof}
Let $\ell$ be the length of $\sigma$ and define the function
$v \colon \sigma \times K \to \C$ with
\[
v(\xi, z) = \frac{h(\xi)}{\xi - z}.
\]
It is clearly continuous, therefore it is uniformly continuous. In
particular, there exists a $\delta > 0$ such that
\[
\abs{v(\xi, z) - v(\xi', z)} < \frac{\varepsilon}{\ell}
\]
for all $z \in \C$ and $\abs{\xi - \xi'} < \delta$.

Let $\tau_1, \dots, \tau_m$ be the partition of $\sigma$ into line
segments of length $d < \sigma$. Choose points $w_k \in \tau_k$ and
set $c_k = - h(w_k) d$. We therefore have
\[
\abs{\lint_{\tau_k} v(\xi, z)\,d\xi - \frac{c_k}{z - w_k}} =
\abs{\lint_{\tau_k} v(\xi, z)\,d\xi - d \cdot v(w_k, z)} <
d \cdot \frac{\varepsilon}{\ell}.
\]
Summing up, we get the desired inequality.
\end{proof}

\begin{lema}
Let $\Omega \subseteq \C$ be an open set and $K \subseteq \Omega$
be a non-empty compact. Then there exist finitely many line
segments $\sigma_1, \dots, \sigma_n$ in $\Omega \setminus K$ such
that for any holomorphic function $f \in \mathcal{O}(\Omega)$ and
$\varepsilon > 0$ there exists a rational function $q$ of the form
\[
q(z) = \sum_{k=1}^m \frac{c_k}{z - w_k},
\]
where $c_k \in \C$ and $w_k \in \sigma$, such that
\[
\norm{f - q}_K < \varepsilon.
\]
\end{lema}

\begin{proof}
By lemma~\ref{app:lm:seg} there exist line segments $\sigma_k$ such
that
\[
f(z) = \frac{1}{2 \pi i} \sum_{k=1}^n
\lint_{\sigma_k} \frac{f(\xi)}{\xi - z}\,d\xi
\]
for all $z \in K$. Now just apply lemma~\ref{app:lm:app_seg} to
each line segment separately.
\end{proof}

\begin{izrek}[Runge's little theorem]
\index{Runge's little theorem}
Let $K \subseteq \C$ be a compact set. If $\C \setminus K$ is
connected, then for every holomorphic function on a neighbourhood
of $K$ and $\varepsilon > 0$ there exists a polynomial
$F \in \C[z]$ such that
\[
\norm{F - f}_K < \varepsilon.
\]
\end{izrek}
